Phonon polaritonics in two-dimensional materials

### 2019-01-06 paper.tex
minor edits

In this Letter, we develop a theoretical and computational framework with quantitative, concrete examples to enable a new understanding of optical phononics in two-dimensional materials. We do so by deriving a universal form for the conductivity of a polar monolayer,  which depends solely on the LO and TO phonon frequencies---and their dispersion with momentum---in the two-dimensional system. Using parameters from Ref.~\citenum{sohier2017breakdown} for the canonical two-dimensional polar monolayer --- hexagonal boron nitride (hBN) --- we present the confinement and propagation losses of the 2D phononic modes, identifying the frequency region where they should be most easily detected. Finally, we show that these modes enable extreme light-matter interaction between emitters and polar materials, showing that for atom-like emitters, their spontaneous decay can be enhanced by up to eight orders of magnitude by the emitter--LO phonon coupling. For an infrared emitter with a sufficiently high free-space radiative decay rate ($> 10^6 s$ $^{-1}$), the enhanced linewidth of the emitter becomes comparable to the spacing between different phonon polaritonic resonances of an hBN nanostructure. This suggests the possibility of realizing the multi-mode strong coupling and ultra-strong coupling regimes of quantum electrodynamics in a two-dimensional hBN platform. Our results for hBN are particularly relevant due its widespread use in 2D van der Waals heterostructures. In addition to providing functionality as a layer which improves the electrical and optical properties of other 2D materials, such as graphene, our results suggest that in these heterostructures, the hBN layers could provide a platform for nanophotonics and quantum optics which operates in the mid-infrared spectral range.

@@

In Figure~\ref{fig:3}, we consider the coupling of a dipole emitter to localized phonon-polaritons of nanostructured monolayer hBN. For simplicity, we consider hBN nanostructured as a disk, which leads to the formation of sharp resonances quantized along the azimuthal and radial directions. The disk is taken to have a radius of \SI{20}{cm} and a loss rate $\tau^{-1} = 5$ cm$^{-1}$, which is of the order of the loss rate in bulk hBN. In Figure~\ref{fig:3}, we plot the enhancement of the spontaneous emission rate for an external emitter polarized perpendicularly to the plane of the disk and placed \SI{5}{\nm} away from the surface of the disk. We find that the rate of spontaneous emission of 2D optical phonons is approximately 8 orders of magnitude larger than the rate of spontaneous emission of photons in the far field. For an infrared emitter at a transition wavelength of \SI{7}{\micro\m} with a free-space radiative lifetime of \SI{1}{\micro\s}, \SI{5}{\nm} away from an hBN disk, the coupling rate to 2D optical phonons (about $\SI{1e14}{\per\s}$ or 0.065 eV) would be on the same scale as the optical phonon frequency itself (about $1430$ cm$^{-1}$ or $\SI{2.70e14}{\per\s}$ or 0.18 eV). This coupling rate thus implies coupling between an emitter and the field in the regime of ultra-strong coupling. Moreover, the coupling rate for the 20 nm disk (orange), for an emitter with a far field decay rate of $> \SI{3e14}{\per\s}$ would have a sufficient coupling strength to the resonances in orange for its linewidth to span multiple resonances and thus be in a multi-mode ultra-strong coupling regime. Thus, the extreme confinement of electromagnetic energy associated with LO phonons in two dimensions enables the possibility of realizing ultrastrong coupling of an atom or molecule with optical phonons in a polar material, allowing the potential realization of new coupled states of quantum emitters and phonons such as atom-phonon polariton bound states.

>>>>

In this Letter, we develop a theoretical and computational framework with quantitative, concrete examples to enable a new understanding of optical phononics in two-dimensional materials. We do so by deriving a universal form for the conductivity of a polar monolayer,  which depends solely on the LO and TO phonon frequencies---and their dispersion with momentum---in the two-dimensional system. Using parameters from Ref.~\citenum{sohier2017breakdown} for the canonical two-dimensional polar monolayer --- hexagonal boron nitride (hBN) --- we present the confinement and propagation losses of the 2D phononic modes, identifying the frequency region where they should be most easily detected. Finally, we show that these modes enable extreme light-matter interaction between emitters and polar materials, showing that for atom-like emitters, their spontaneous decay can be enhanced by up to eight orders of magnitude by the emitter--LO phonon coupling. For an infrared emitter with a sufficiently high free-space radiative decay rate ($> 10^6 s$ $^{-1}$), the enhanced linewidth of the emitter becomes comparable to the spacing between different phonon polaritonic resonances of an hBN nanostructure. This suggests the possibility of realizing the multi-mode strong coupling and ultra-strong coupling regimes of quantum electrodynamics in a two-dimensional hBN platform. Our results for hBN are particularly relevant due its widespread use in 2D van der Waals heterostructures. In addition to providing functionality as a layer which improves the electrical and optical properties of other 2D materials, such as graphene, our results suggest that in these heterostructures, the hBN layers could provide a platform for nanophotonics and quantum optics which operates in the mid-infrared spectral range. While we focus on hBN in this manuscript, many of our findings apply to other polar monolayers.

@@

In Figure~\ref{fig:3}, we consider the coupling of a dipole emitter to localized phonon-polaritons of nanostructured monolayer hBN. For simplicity, we consider hBN nanostructured as a disk, which leads to the formation of sharp resonances quantized along the azimuthal and radial directions. The disk is taken to have a radius of \SI{20}{cm} and a loss rate $\tau^{-1} = 5$ cm$^{-1}$, which is of the order of the loss rate in bulk hBN. We parameterize the coupling between the dipole and phonon-polaritons through the rate of spontaneous emission $\Gamma$ of phonon polaritons by the dipole. It is related to the Green's function of the Maxwell equations for the nanostructure via the expression \cite{novotny2012principles}:
\begin{equation}
\frac{\Gamma}{\Gamma_0} = 1 + \frac{6\pi c}{\omega_0}\hat{n}\cdot\text{Im }G(\mathbf{r}_0,\mathbf{r}_0,\omega_0)\cdot\hat{n},
\end{equation}
where $\hat{n}$ is the orientation of the dipole, $\mathbf{r}_0$ is its position, and $\omega_0$ is its frequency. The Green's function is computed through a boundary element method (as in Ref. \cite{jin2017infrared}) assuming the validity of the electrostatic approximation, which is well-respected due to the extreme sub-wavelength confinement of the electromagnetic field of the phonon polariton. In Figure~\ref{fig:3}, we plot the enhancement of the spontaneous emission rate for an external emitter polarized perpendicularly to the plane of the disk and placed \SI{5}{\nm} away from the surface of the disk. We find that the rate of spontaneous emission of 2D optical phonons is approximately 8 orders of magnitude larger than the rate of spontaneous emission of photons in the far field. We note that in this approach, the coupling of the dipole to phonon polaritons is manifested through the phonon contribution to the dielectric function of the disk. This should be equivalent to an approach that considers the coupling of a bound electron in an emitter to LO phonons in the disk through a (two-dimensional) Frohlich coupling -- i.e., a coupling of the atomic electron to the electric potential resulting from the polarization associated with an LO phonon mode \cite{sohier2016two}.

In \ref{fig:3}, we show that for an infrared emitter at a transition wavelength of \SI{7}{\micro\m} with a free-space radiative lifetime of \SI{1}{\micro\s}, \SI{5}{\nm} away from an hBN disk, the coupling rate to 2D optical phonons (about $\SI{1e14}{\per\s}$ or 0.065 eV) would be on the same scale as the optical phonon frequency itself (about $1430$ cm$^{-1}$ or $\SI{2.70e14}{\per\s}$ or 0.18 eV). This coupling rate thus implies coupling between an emitter and the field in the regime of ultra-strong coupling. Moreover, the coupling rate for the 20 nm disk (orange), for an emitter with a far field decay rate of $> \SI{3e14}{\per\s}$ would have a sufficient coupling strength to the resonances in orange for its linewidth to span multiple resonances and thus be in a multi-mode ultra-strong coupling regime. Thus, the extreme confinement of electromagnetic energy associated with LO phonons in two dimensions enables the possibility of realizing ultrastrong coupling of an atom or molecule with optical phonons in a polar material, allowing the potential realization of new coupled states of quantum emitters and phonons such as atom-phonon polariton bound states.
